---
title: Exploratory data analysis and quality control
vignette: >
  % \VignetteIndexEntry{Quality control}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r setup, message = FALSE}
library(scater)
library(MouseGastrulationData)
```

Let's start with an example dataset.

```{r}
sce <- WTChimeraData(samples=5, type="raw")
sce <- sce[[1]]
dim(sce) ## 29,453 522,554
```
Delete cols and rows which have only zero values.

```{r}
sce <- sce[rowSums(assay(sce)) > 0,]
sce <- sce[,colSums(assay(sce)) > 0]
dim(sce) ## 19,852 515,529
```

# Droplet processing

Sections 7.2.1 and 7.2.2 of OSCA advanced.

# Quality control

Sections 1.2, 1.3.2, 1.4, 1.5 of OSCA basics.

```{r}
is.mito <- grep(“mt”, rowData(sce)[,“SYMBOL”])

library(scuttle)
df <- perCellQCMetrics(sce, subsets=list(Mito=is.mito),
                       BPPARAM=BiocParallel::MulticoreParam(3))

low.lib <- isOutlier(df$sum, type=“lower”, log=TRUE)
summary(low.lib)

high.mito <- isOutlier(df$sum, type=“higher”)
summary(high.mito)

reasons <- perCellQCFilters(df, sub.fields=“subsets_Mito_percent”)
summary(reasons$discard)

## Diagnosing cell type loss
lost <- calculateAverage(sce[,reasons$discard])
kept <- calculateAverage(sce[,!reasons$discard])

library(edgeR)
logged <- cpm(cbind(lost, kept), log=TRUE, prior.count=2)
logFC <- logged[,1] - logged[,2]
abundance <- rowMeans(logged)
plot(abundance, logFC, xlab=“Average count”, ylab=“Log-FC (lost/kept)“, pch=16)
points(abundance[is.mito], logFC[is.mito], col=“dodgerblue”, pch=16)
                       
```

# Normalization

Sections 2.1, 2.2, 2.3, 2.5, of OSCA basics.

```{r}
sce <- logNormCounts(sce)
```

# Feature Selection

Sections 3.1, 3.2, 3.5 of OSCA basics.

## 3.2 Quantifying per-gene variation (very long to run)
```{r}
fit.sce <- metadata(dec.sce)
plot(fit.sce$mean, fit.sce$var, xlab = “Mean of log-expression”,
     ylab = “Variance of log-expression”)
curve(fit.sce$trend(x), col = “dodgerblue”, add = TRUE, lwd = 2)

dec.sce[order(dec.sce$bio, decreasing=FALSE),]
```

## 3.5 Selecting highly variable genes
```{r}
library(scran)
hvg.sce.var <- getTopHVGs(sce, n=1000)
str(hvg.sce.var)

sce.new <- sce[hvg.sce.var, ]

dim(sce) # 19852 515529
dim(sce.new) # 1000 515529
```


# Dimensionality Reduction

All sections of Chapter 4 of OSCA basics.

## Principal Component Analysis (PCA)
```{r}
sce <- BiocSingular::runPCA(sce.new, 50, BPPARAM=BiocParallel::MulticoreParam(2), BSPARAM=BiocSingular::RandomParam())

plotPCA(sce, colour_by=I(librarySizeFactors(sce)))
reducedDimNames(sce)
dim(reducedDim(sce, “PCA”))

percent.var <- attr(reducedDim(sce), “percentVar”)
plot(percent.var, log=“y”, xlab=“PC”, ylab=“Variance explained (%)“)

library(scater)
plotReducedDim(sce, dimred=“PCA”)
plotReducedDim(sce, dimred=“PCA”, ncomponents=3)
```

## t-stochastic neighbor embedding (t-SNE) (very long to run)
```{r}
set.seed(100)
sce <- runTSNE(sce, dimred=“PCA”, perplexity=5)
out5 <- plotReducedDim(sce, dimred=“TSNE”) + ggtitle(“perplexity = 5")

set.seed(100)
sce <- runTSNE(sce, dimred=“PCA”, perplexity=20)
out20 <- plotReducedDim(sce, dimred=“TSNE”) + ggtitle(“perplexity = 20”)

set.seed(100)
sce <- runTSNE(sce, dimred=“PCA”, perplexity=80)
out80 <- plotReducedDim(sce, dimred=“TSNE”) + ggtitle(“perplexity = 80")

gridExtra::grid.arrange(out5, out20, out80, ncol=3)
```

## Uniform manifold approximation and projection (UMAP) (very long to run)
```{r}
set.seed(1100101001)
sce <- runUMAP(sce, dimred=“PCA”, BPPARAM=BiocParallel::MulticoreParam(3))
plotReducedDim(sce, dimred=“UMAP”)
```

# Doublet identification

Sections 8.1, 8.3 of OSCA advanced.

## Computing doublet desities (very long to run)
```{r}
library(BiocSingular)
library(scDblFinder)
set.seed(100)

dbl.dens <- computeDoubletDensity(sce, subset.row=hvg.sce.var,
                                  d=ncol(reducedDim(sce)))
summary(dbl.dens)
sce$DoubletScore <- dbl.dens

#plotTSNE(sce, colour_by=“DoubletScore”) #very long to run

dbl.calls <- doubletThresholding(data.frame(score=dbl.dens),
                                 method=“griffiths”, returnType=“call”)
summary(dbl.calls)

plotColData(sce, x=“label”, y=“DoubletScore”, colour_by=I(dbl.calls))
```

## Doublet classification (very long to run)
```{r}
set.seed(10010101)
sce.dbl <- scDblFinder(sce, clusters=colLabels(sce))
plotTSNE(sce.dbl, colour_by=“scDblFinder.score”)
table(sce.dbl$scDblFinder.class)
```

