---
title: Working with large data
vignette: >
  % \VignetteIndexEntry{Working with large data}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

```{r setup, include=FALSE}
rm(list = ls())
gc()
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(BiocStyle)
```

# Motivation 

Advances in scRNA-seq technologies have increased the number of cells that can 
be assayed in routine experiments.
Public databases such as [GEO](https://www.ncbi.nlm.nih.gov/geo/) are continually
expanding with more scRNA-seq studies, 
while large-scale projects such as the
[Human Cell Atlas](https://www.humancellatlas.org/) are expected to generate
data for billions of cells.
For effective data analysis, the computational methods need to scale with the
increasing size of scRNA-seq data sets.
This section discusses how we can use various aspects of the Bioconductor 
ecosystem to tune our analysis pipelines for greater speed and efficiency.

# Out of memory representations

The count matrix is the central structure around which our analyses are based.
In most of the previous chapters, this has been held fully in memory as a dense 
`matrix` or as a sparse `dgCMatrix`.
Howevever, in-memory representations may not be feasible for very large data sets,
especially on machines with limited memory.
For example, the 1.3 million brain cell data set from 10X Genomics 
([Zheng et al., 2017](https://doi.org/10.1038/ncomms14049))
would require over 100 GB of RAM to hold as a `matrix` and around 30 GB as a `dgCMatrix`.
This makes it challenging to explore the data on anything less than a HPC system.

The obvious solution is to use a file-backed matrix representation where the 
data are held on disk and subsets are retrieved into memory as requested.
While a number of implementations of file-backed matrices are available 
(e.g.,
[bigmemory](https://cran.r-project.org/web/packages/bigmemory/index.html), 
[matter](https://bioconductor.org/packages/matter)), 
we will be using the implementation from the 
[HDF5Array](https://bioconductor.org/packages/HDF5Array) package.
This uses the popular HDF5 format as the underlying data store, which provides
a measure of standardization and portability across systems.
We demonstrate with a subset of 20,000 cells from the 1.3 million brain cell 
data set, as provided by the
[TENxBrainData](https://bioconductor.org/packages/TENxBrainData) package.

```{r tenx-brain, message = FALSE}
library(TENxBrainData)
sce.brain <- TENxBrainData20k() 
sce.brain
```

Examination of the `SingleCellExperiment` object indicates that the count matrix
is a `HDF5Matrix`.
From a comparison of the memory usage, it is clear that this matrix object is
simply a stub that points to the much larger HDF5 file that actually contains
the data.
This avoids the need for large RAM availability during analyses.

```{r tenx-brain-size}
counts(sce.brain)
object.size(counts(sce.brain))
file.info(path(counts(sce.brain)))$size
```

Manipulation of the count matrix will generally result in the creation of a
`DelayedArray` object from the 
[DelayedArray](https://bioconductor.org/packages/DelayedArray) package.
This remembers the operations to be applied to the counts and stores them in
the object, to be executed when the modified matrix values are realized for use
in calculations.
The use of delayed operations avoids the need to write the modified values to a
new file at every operation, which would unnecessarily require time-consuming disk I/O.

```{r}
tmp <- counts(sce.brain)
tmp <- log2(tmp + 1)
tmp
```

Many functions described in the previous workflows are capable of accepting 
`HDF5Matrix` objects.
This is powered by the availability of common methods for all matrix
representations (e.g., subsetting, combining, methods from 
[DelayedMatrixStats](https://bioconductor.org/packages/DelayedMatrixStats) 
as well as representation-agnostic C++ code 
using [beachmat](https://bioconductor.org/packages/beachmat).
For example, we compute QC metrics below with the same `calculateQCMetrics()` 
function that we used in the other workflows.

```{r}
library(scater)
is.mito <- grepl("^mt-", rowData(sce.brain)$Symbol)
qcstats <- perCellQCMetrics(sce.brain, subsets = list(Mt = is.mito))
qcstats
```

Needless to say, data access from file-backed representations is slower than that
from in-memory representations.
The time spent retrieving data from disk is an unavoidable cost of reducing
memory usage.
Whether this is tolerable depends on the application.
One example usage pattern involves performing the heavy computing quickly with 
in-memory representations on HPC systems with plentiful memory, and then
distributing file-backed counterparts to individual users for exploration and 
visualization on their personal machines.

# Parallelization

Parallelization of calculations across genes or cells is an obvious strategy for
speeding up scRNA-seq analysis workflows.

The `r Biocpkg("BiocParallel")` package provides a common interface for parallel
computing throughout the Bioconductor ecosystem, manifesting as a `BPPARAM`
argument in compatible functions. We can also use `BiocParallel` with more
expressive functions directly through the package's interface.

### Basic use

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(BiocParallel)
```

`BiocParallel` makes it quite easy to iterate over a vector and distribute the
computation across workers using the `bplapply` function. Basic knowledge
of `lapply` is required.

In this example, we find the square root of a vector of numbers in parallel
by indicating the `BPPARAM` argument in `bplapply`.

```{r}
param <- MulticoreParam(workers = 1)
bplapply(
    X = c(4, 9, 16, 25),
    FUN = function(x) { sqrt(x) },
    BPPARAM = param
)
```

**Note**. The number of workers is set to 1 due to continuous testing resource
limitations.

There exists a diverse set of parallelization backends depending on available
hardware and operating systems.

For example, we might use forking across two cores to parallelize the variance
calculations on a Unix system:

```{r parallel-mc, message = FALSE}
library(MouseGastrulationData)
library(scran)
sce <- WTChimeraData(samples = 5, type = "processed")
sce <- logNormCounts(sce)
dec.mc <- modelGeneVar(sce, BPPARAM = MulticoreParam(2))
dec.mc
```

Another approach would be to distribute jobs across a network of computers,
which yields the same result:

```{r parallel-snow, eval=FALSE}
dec.snow <- modelGeneVar(sce, BPPARAM = SnowParam(2))
```

For high-performance computing (HPC) systems with a cluster of compute nodes, 
we can distribute jobs via the job scheduler using the `BatchtoolsParam` class.
The example below assumes a SLURM cluster, though the settings can be easily 
configured for a particular system 
(see `r Biocpkg("BiocParallel", "BiocParallel_BatchtoolsParam.pdf", "here")` for
details).

```{r batch-param, eval=FALSE}
# 2 hours, 8 GB, 1 CPU per task, for 10 tasks.
rs <- list(walltime = 7200, memory = 8000, ncpus = 1)
bpp <- BatchtoolsParam(10, cluster = "slurm", resources = rs)
```

Parallelization is best suited for CPU-intensive calculations where the division
of labor results in a concomitant reduction in compute time. It is not suited
for tasks that are bounded by other compute resources, e.g., memory or file I/O
(though the latter is less of an issue on HPC systems with parallel read/write).
In particular, R itself is inherently single-core, so many of the
parallelization backends involve (i) setting up one or more separate R sessions,
(ii) loading the relevant packages and (iii) transmitting the data to that
session. Depending on the nature and size of the task, this overhead may
outweigh any benefit from parallel computing.

# Fast approximations

## Nearest neighbor searching

Identification of neighbouring cells in PC or expression space is a common procedure
that is used in many functions, e.g., `buildSNNGraph()`, `doubletCells()`.
The default is to favour accuracy over speed by using an exact nearest neighbour
(NN) search, implemented with the $k$-means for $k$-nearest neighbours algorithm.
However, for large data sets, it may be preferable to use a faster approximate 
approach.

The `r Biocpkg("BiocNeighbors")` framework makes it easy to switch between search
options by simply changing the `BNPARAM` argument in compatible functions.
To demonstrate, we will use the wild-type chimera data for which we had applied
graph-based clustering using the Louvain algorithm for community detection:

```{r pca-and-cluster}
library(bluster)
sce <- runPCA(sce)
colLabels(sce) <- clusterCells(sce, use.dimred = "PCA",
                               BLUSPARAM = NNGraphParam(cluster.fun = "louvain"))
```

The above clusters on a nearest neighbor graph generated with
an exact neighbour search.
We repeat this below using an approximate search, implemented using the
[Annoy](https://github.com/spotify/Annoy) algorithm.
This involves constructing a `AnnoyParam` object to specify the search algorithm
and then passing it to the parameterization of the `NNGraphParam()` function.
The results from the exact and approximate searches are consistent with most
clusters from the former re-appearing in the latter.
This suggests that the inaccuracy from the approximation can be largely ignored.

```{r cluster-annoy}
library(scran)
library(BiocNeighbors)
clusters <- clusterCells(sce, use.dimred = "PCA",
                         BLUSPARAM = NNGraphParam(cluster.fun = "louvain",
                                                  BNPARAM = AnnoyParam()))
table(exact = colLabels(sce), approx = clusters)
```

```{r check-annoy, echo = FALSE}
# Check that clusterings are very similar
library(bluster)
rand <- pairwiseRand(colLabels(sce), clusters, mode = "index")
stopifnot(rand > 0.85)
```

Note that Annoy writes the NN index to disk prior to performing the search.
Thus, it may not actually be faster than the default exact algorithm for small
datasets, depending on whether the overhead of disk write is offset by the
computational complexity of the search.
It is also not difficult to find situations where the approximation deteriorates,
especially at high dimensions, though this may not have an appreciable impact on
the biological conclusions.

```{r bad-approx}
set.seed(1000)
y1 <- matrix(rnorm(50000), nrow = 1000)
y2 <- matrix(rnorm(50000), nrow = 1000)
Y <- rbind(y1, y2)
exact <- findKNN(Y, k = 20)
approx <- findKNN(Y, k = 20, BNPARAM = AnnoyParam())
mean(exact$index != approx$index)
```

## Singular value decomposition 

The singular value decomposition (SVD) underlies the PCA used throughout our
analyses, e.g., in `denoisePCA()`, `fastMNN()`, `doubletCells()`.
(Briefly, the right singular vectors are the eigenvectors of the gene-gene
covariance matrix, where each eigenvector represents the axis of maximum remaining
variation in the PCA.)
The default `base::svd()` function performs an exact SVD that is not performant
for large datasets.
Instead, we use fast approximate methods from the `r CRANpkg("irlba")` and
`r CRANpkg("rsvd")` packages, conveniently wrapped into the 
`r Biocpkg("BiocSingular")` package for ease of use and package development.
Specifically, we can change the SVD algorithm used in any of these functions by
simply specifying an alternative value for the `BSPARAM` argument.

```{r fast-svd}
library(scater)
library(BiocSingular)

# As the name suggests, it is random, so we need to set the seed.
set.seed(101000)
r.out <- runPCA(sce, ncomponents = 20, BSPARAM = RandomParam())
str(reducedDim(r.out, "PCA"))

set.seed(101001)
i.out <- runPCA(sce, ncomponents = 20, BSPARAM = IrlbaParam())
str(reducedDim(i.out, "PCA"))
```

Both IRLBA and randomized SVD (RSVD) are much faster than the exact SVD with
negligible loss of accuracy.
This motivates their default use in many `r Biocpkg("scran")` and `r Biocpkg("scater")`
functions, at the cost of requiring users to set the seed to guarantee reproducibility.
IRLBA can occasionally fail to converge and require more iterations
(passed via `maxit=` in `IrlbaParam()`), while RSVD involves an explicit trade-off
between accuracy and speed based on its oversampling parameter (`p=`) and number
of power iterations (`q=`).
We tend to prefer IRLBA as its default behavior is more accurate, though RSVD is
much faster for file-backed matrices.

# Interoperability with popular single-cell analysis ecosytems

## Seurat

[Seurat](https://satijalab.org/seurat) is an R package designed for QC, analysis,
and exploration of single-cell RNA-seq data. Seurat can be used to identify and
interpret sources of heterogeneity from single-cell transcriptomic measurements,
and to integrate diverse types of single-cell data. Seurat is developed and
maintained by the [Satija lab](https://satijalab.org/seurat/authors.html)
and is released under the [MIT license](https://opensource.org/license/mit/).

```{r load-seurat, message = FALSE}
library(Seurat)
```

Although the basic processing of single-cell data with Bioconductor packages
(described in the [OSCA book](https://bioconductor.org/books/release/OSCA/))
and with Seurat is very similar and will produce overall roughly identical
results, there is also complementary functionality with regard to cell type
annotation, dataset integration, and downstream analysis.
To make the most of both ecosystems it is therefore beneficial to be able 
to easily switch between a `SeuratObject` and a `SingleCellExperiment`.
See also the Seurat
[conversion vignette](https://satijalab.org/seurat/articles/conversion_vignette.html)
for conversion to/from other popular single cell formats such as the AnnData
format used by [scanpy](https://scanpy.readthedocs.io/en/stable/).

Here, we demonstrate converting the Seurat object produced in Seurat's
[PBMC tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
to a `SingleCellExperiment` for further analysis with functionality from
OSCA/Bioconductor.
We therefore need to first install the 
[SeuratData](https://github.com/satijalab/seurat-data) package, which is available
from GitHub only. 

```{r, eval = FALSE}
BiocManager::install("satijalab/seurat-data")
```

We then proceed by loading all required packages and installing the PBMC dataset:

```{r seurat-data}
library(SeuratData)
InstallData("pbmc3k")
```

We then load the dataset as an `SeuratObject` and convert it to a 
`SingleCellExperiment`.

```{r seurat_singlecell}
# Use PBMC3K from SeuratData
pbmc <- LoadData(ds = "pbmc3k", type = "pbmc3k.final")
pbmc
pbmc.sce <- as.SingleCellExperiment(pbmc)
pbmc.sce
```

Seurat also allows conversion from `SingleCellExperiment` objects to Seurat objects; 
we demonstrate this here on the wild-type chimera mouse gastrulation dataset. 

```{r sce2sobj, message = FALSE}
sce <- WTChimeraData(samples = 5, type = "processed")
assay(sce) <- as.matrix(assay(sce))
sce <- logNormCounts(sce)
sce
```

After some processing of the dataset, the actual conversion is carried out with
the `as.Seurat` function.

```{r, warning = FALSE}
sobj <- as.Seurat(sce)
Idents(sobj) <- "celltype.mapped"
sobj
```

## Scanpy

[Scanpy](https://scanpy.readthedocs.io) is a scalable toolkit for analyzing
single-cell gene expression data built jointly with
[anndata](https://anndata.readthedocs.io/). It includes
preprocessing, visualization, clustering, trajectory inference and differential
expression testing. The Python-based implementation efficiently deals with
datasets of more than one million cells. Scanpy is developed and maintained by
the [Theis lab]() and is released under a
[BSD-3-Clause license](https://github.com/scverse/scanpy/blob/master/LICENSE).
Scanpy is part of the [scverse](https://scverse.org/), a Python-based ecosystem
for single-cell omics data analysis.

At the core of scanpy's single-cell functionality is the `anndata` data structure,
scanpy's integrated single-cell data container, which is conceptually very similar
to Bioconductor's `SingleCellExperiment` class.

Bioconductor's `r Biocpkg("zellkonverter")` package provides a lightweight
interface between the Bioconductor `SingleCellExperiment` data structure and the
Python `AnnData`-based single-cell analysis environment. The idea is to enable
users and developers to easily move data between these frameworks to construct a
multi-language analysis pipeline across R/Bioconductor and Python.

```{r load-zellkonv, message = FALSE}
library(zellkonverter)
```

The `readH5AD()` function can be used to read a `SingleCellExperiment` from an
H5AD file. Here, we use an example H5AD file contained in the  `r Biocpkg("zellkonverter")`
package.

```{r read-h5ad}
example_h5ad <- system.file("extdata", "krumsiek11.h5ad",
                            package = "zellkonverter")
readH5AD(example_h5ad)
```

We can also write a `SingleCellExperiment` to an H5AD file with the
`writeH5AD()` function. This is demonstrated below on the wild-type
chimera mouse gastrulation dataset. 

```{r write-h5ad}
out.file <- tempfile(pattern = ".h5ad")
writeH5AD(sce, file = out.file)
```

The resulting H5AD file can then be read into Python using scanpy's
[read_h5ad](https://scanpy.readthedocs.io/en/stable/generated/scanpy.read_h5ad.html)
function and then directly used in compatible Python-based analysis frameworks.

# Session Info

```{r sessionInfo}
sessionInfo()
```

# Further Reading

* OSCA book, [Chapter 14](https://bioconductor.org/books/release/OSCA.advanced/dealing-with-big-data.html): Dealing with big data 
* The `BiocParallel` `r Biocpkg("BiocParallel", vignette = "Introduction_To_BiocParallel.html", label = "intro vignette")`. 

# Exercises

1. Out of memory representation: Write the counts matrix of the wild-type chimera
mouse gastrulation dataset to an HDF5 file. Create another counts matrix that
reads the data from the HDF5 file. Compare memory usage of holding the entire
matrix in memory as opposed to holding the data out of memory.  

Hint: see the `HDF5Array` function for reading from HDF5 and the `writeHDF5Array`
function for writing to HDF5 from the `r Biocpkg("HDF5Array")` package.

2. Parallelization: Perform a PCA analysis of the  wild-type chimera mouse gastrulation
dataset using a multicore backend for parallel computation. Compare the runtime
of performing the PCA either in serial execution mode, in multicore execution
mode with 2 workers, and in multicore execution mode with 3 workers.

Hint: use the function `system.time` to obtain the runtime of each job.

3. Conversion to Seurat: The [scRNAseq](https://bioconductor.org/packages/scRNAseq)
package provides gene-level counts for a collection of public scRNA-seq datasets,
stored as `SingleCellExperiment` objects with annotated cell- and gene-level metadata.
Consult the vignette of the [scRNAseq](https://bioconductor.org/packages/scRNAseq)
package to inspect all available datasets and select a dataset of your choice.
Convert the chosen dataset to a Seurat object and produce a PCA plot with cells
colored by a cell metadata column of your choice.

Hint: use Seurat's `DimPlot` function.
